Three distinct patterns of mental health response following accidents in mountain sports – a follow-up study of individuals treated at a tertiary trauma center

### 2023-11-07 paper/markdown/manuscript.Rmd
Feedback, October 2023, markdown and figures

author: 
  - Hanna Salvotti, MD:
      institute: [psych, regensburg]
  - Piotr Tymoszuk, PhD:
      institute: [daas]
  - Mathias Ströhle, MD:
      institute: [kufstein]
  - Peter Paal, MD: 
      institute: [salzburg, kurasi, himed, icar]
  - Hermann Brugger, MD: 
      institute: [eurac, ismm]
  - Martin Faulhaber, PhD: 
      institute: [sport]
  - Nicola Kruijer, MD: 
      institute: [trauma]
  - Barbara Sperner-Unterweger, MD: 
      institute: [psych]
  - Katharina Hüfner, MD:
      institute: [psych]
      email: katharina.huefner@tirol-kliniken.at
      correspondence: true

institute:
  - psych: Department of Psychiatry, Psychotherapy, Psychosomatics and Medical Psychology, University Hospital for Psychiatry II, Medical University of Innsbruck, Innsbruck, Austria.
  - regensburg: Department of Neurosurgery, University Hospital of Regensburg, Regensburg, Germany
  - daas: Data Analysis as a Service Tirol, Innsbruck, Austria
  - kufstein: Department of Anesthesiology and Critical Care Medicine, Bezirkskrankenhaus Kufstein, Kufstein, Austria.
  - salzburg: Department of Anesthesiology and Intensive Care Medicine, Paracelsus Medical University, Salzburg, Austria.
  - kurasi:  Austrian Board of Mountain Safety (Österreichisches Kuratorium fur Alpine Sicherheit), Innsbruck, Austria.
  - himed: Austrian Society of Mountain and High-Altitude Medicine, Mieming, Austria. 
  - icar: International Commission for Mountain Emergency Medicine (ICAR), Kloten, Switzerland.
  - eurac: EURAC Research, Institute for Mountain Emergency Medicine, Bolzano/Bozen, Italy.
  - ismm: International Society of Mountain Medicine (ISMM).
  - sport: Department of Sport Science, University of Innsbruck, Innsbruck, Austria.
  - trauma: Department of Orthopedics and Traumatology, Medical University of Innsbruck, Innsbruck, Austria.

@@

__Background:__ Mountain sports are inherently linked to physical injury risk. 
We aimed to characterize mental health following mountain sport accidents requiring professional medical management.

__Methods:__ Adult victims of mountain sport accidents treated at the Innsbruck University Hospital (Austria) between 2008 and 2020 (n = `r nrow(ptsd$dataset)`) participated in a cross-sectional survey study at median `r round(median(ptsd$dataset$obs_time)/30.436)` months after the trauma center admission. Symptoms of anxiety, depression, somatization and panic disorder (PHQ), resilience (RS-13), sense of coherence (SOC-9L), quality of life (EUROHIS-QOL), post-traumatic growth (PTGI) and post-traumatic stress disorder (PTSD, PCL-5), sociodemographic and clinical information were obtained from an on-line survey and electronic health records. 
Mental health patterns were investigated by semi-supervised medoid clustering and modeled by machine learning.

__Results:__ At least one diagnostic criterion of PTSD was met by `r signif(table(ptsd$dataset$dsm5_cluster_class)[2]/nrow(ptsd$dataset), 2) * 100`% of participants. 
Three comparably sized subsets of patients were identified: (1) neutral, (2) post-traumatic growth and (3) post-traumatic stress cluster. The post-traumatic stress cluster was characterized by frequent mental disorder symptoms, and low resilience, sense of coherence and quality of life as well as by low age, the highest frequency of pre-existing mental disorders and common physical health consequences of the accident. Reliability of machine learning-based prediction of the cluster assignment based on `r length(class_globals$variables$early)` variables available during acute treatment was limited.

__Conclusion:__ A subset of mountain sport accident victims shows symptoms of mental health disorders including PTSD. 
Since early identification of these vulnerable patients remains challenging, low-threshold psychological support is key for a successful interdisciplinary management of mountain sport accidents.

@@

mountain sport accident, mental health, post-traumatic stress disorder, post-traumatic growth, machine learning

@@

Physical activity in mountains, such as hiking, climbing, alpine skiing or ski touring, gain popularity, a trend that has been accelerated by the COVID-19 pandemic [@Hedenborg2022]. 
Outdoor physical activity in general and mountain sport in particular exerts beneficial effects on mental health [@Ower2019; @Niedermeier2017a; @Niedermeier2017]. 
However, mountain sports are also associated with the risk of injury [@Faulhaber2017; @Faulhaber2020], which could foster development of mental disorders including post-traumatic stress disorder (PTSD).

PTSD symptoms include hyperarousal, panic, somatic symptoms, sleep disorders and unwilling recalls of the trauma (flashbacks) [@Bryant2019; @AmericanPsychiatricAssociation2013; @Howie2019], which can occur following self-experiencing or witnessing of a traumatic event such as actual or threatened death, injury, physical or sexual violence. 
PTSD may co-occur with symptoms of depression and anxiety and substance abuse disorders [@Gagnon-Sanschagrin2022; @Brady2000]. 
Although most individuals have experienced at least one traumatic event during the lifetime [@Benjet2016; @Breslau2004], the prevalence of PTSD in the general European population has been estimated for 0.38 to 6.67% [@Trautmann2018; @Burri2014]. 
Yet, up to 26% of acute physical trauma patients have been reported to develop PTSD [@Joseph2020].

So far it has proven difficult to identify early predictors of subsequent PTSD development in physical trauma patients. 
While some recent studies on specific populations have identified risk factors such as pre-existing psychological distress [@Ziobrowski2021] or symptoms of mental disorders [@Papini2018], others have focused on the arousal and stress reaction during acute trauma and peri-traumatic period as candidate predictors [@Schultebraucks2020]. 
These results require validation and further generalization before they can be implemented into routine care [@Wafa2019; @Shalev2019]. 
Many trauma patients at risk of PTSD undergo intensive acute medical management at emergency departments, which poses a challenge for implementation of early diagnostic and psychological preventive measures [@Linares2017]. 
Furthermore, efficacy of such acute psychological interventions at preventing PTSD in trauma patients is uncertain, probably due to large heterogeneity of study cohorts [@Qi2016].

Post-traumatic growth describes a salutary adjustment psychological reaction to a traumatic experience with strengthened perceptions of self, others, and meaning of the event [@Tedeschi1996]. 
Resilience, i.e. the ability to master adversities has been proposed to bridge traumatic events and post-traumatic growth. 
It was also suggested to explain why most people do not develop PTSD despite the high prevalence of traumatic events in the general population [@Ahmed2007; @Bensimon2012]. 
Overall, mountain sport was associated with high resilience values [@Ower2019].

Reports on mental health following accidents during mountain sports are scarce. 
The available data concern mountain guides, rescuers or avalanche accident survivors [@Leonard2021; @Sommer2004; @Mikutta2022; @Thordardottir2015; @Traber2020]. 
Considering the large number of mountain accident victims, and heterogeneity of accident and injuries, accurate characteristic of mental health of the affected patients is urgently needed. 
In particular, tools allowing for identification of patients at risk of later mental health problems during acute medical management of the mountain accident would facilitate the early targeted psychological support and help to preserve the positive mental health effects of physical activity.

Herein, we analyzed a broad palette of mental health readouts such as anxiety, depression, resilience, sense of coherence, quality of life, post-traumatic growth and signs of PTSD in a collective of `r nrow(ptsd$dataset)` patients treated for injuries during mountain sport accidents at the Medical University of Innsbruck, Austria by means of semi-supervised clustering.  
Additionally, we searched for early sociodemographic and clinical predictors of mental health impairment following a mountain accident with a machine learning approach.

@@

Patients treated for a mountain sport accident at the Department for Orthopedics and Traumatology at the Medical University of Innsbruck between 1^st^ January 2018 and 31^st^ December 2020 (i.e. at least 6 months prior to the start of the study) were screened for participation. 
Patients fulfilling the inclusion criteria: age $\geq$ 18 years, residence in a German-speaking country and proficiency in German (n = `r nrow(ptsd$cleared)`) were invited to participate in the online study survey via conventional mail. 
Out of the invited subjects, `r length(ptsd$interview_ID)` completed the survey and `r nrow(ptsd$dataset)` individuals with the complete psychometry data were analyzed (__Figure \@ref(fig:fig-consort)__, __Supplementary Tables S1__ - __S2__).

@@

Sociodemographic (e.g. age, sex, education, profession, income, prior sport accidents and traumatic events), accident- and recovery-related (e.g. rescue, psychological support, persistent physical health consequences) and psychometric variables were recorded with a cross-sectional on-line survey.

The psychometric battery consisted of German versions of assessment tools for symptoms of 
anxiety (GAD-7: 7-item general anxiety disorder scale) [@Spitzer2006], 
depression (PHQ: patient health questionnaire, PHQ-9, depression module) [@Lowe2002; @Grafe2004], 
panic (PHQ-panic module) [@Lowe2002; @Grafe2004], 
persistent somatic symptoms (PHQ-15, somatization module) [@Kroenke2002], 
resilience (RS13: 13-item resilience scale) [@Leppert2008], 
lack of sense of coherence (SOC-9L: Leipzig 9-item sense of coherence questionnaire) [@Schumacher2000], 
quality of life (EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale) [@Schmidt2006], 
post-traumatic growth (PTGI: post-traumatic growth inventory) [@Tedeschi1996] and 
post-traumatic stress disorder (PCL-5 DSM-5: PTSD checklist for DSM-5) [@Bovin2016]. 
The tools displayed good-to-excellent consistency [@McDonald1999] (McDonald's $\omega$ > 0.8, __Supplementary Table S2__ and __S3__).

Clinically significant symptoms of anxiety (GAD-7 $\geq$ 10), 
depression (PHQ-9 $\geq$ 11) [@Manea2012], 
persistent somatic symptoms (PHQ-15 $\geq$ 11) [@Kroenke2002] and 
resilience classes (low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73)[@Leppert2008] 
were defined with published cutoffs.

Separate scores were calculated for domains B, C, D and E of the PCL-5 tool along with the total score being the sum of all items. 
Each PCL-5 item was scored as 0: not at all, 1: a little bit, 2: moderate, 3: quite a bit and 4: extremely. 
Participants positive for the B domain or C domain PTSD symptoms were identified by at least one item per domain scored with 'moderate' or higher. 
Participants positive for the D or E domain PTSD symptoms were identified by at least two items per domain scored with 'moderate' or higher. 
Significant PTSD symptoms where assumed in participants screened positive for at least one of the B, C, D or E PCL-5 domains [@Bovin2016].

Traumatic events before the accident were assessed with the DIA-X tool [@Maercker2005]. 
Flashbacks frequency was surveyed as none, > 1/year and > 1/month. 
Smoking was surveyed as a single yes/no question. 
Problematic alcohol use was investigated with the CAGE tool [@OBrien2008]. 
Data on the type of the accident mountain sport, injury diagnosis, severity (AIS: abbreviated injury scale) [@Gennarelli2006] and location, hospitalization, surgery and number of surgical ICD-10 diagnoses were extracted from electronic patient's records.

@@

The primary analysis endpoint was assessment of mental health of mountain accident victims by clustering using standardized psychometric tools for symptoms of anxiety, depression, panic, persistent somatic symptoms, quality of life, resilience, sense of coherence, post-traumatic growth and PTSD. 
The secondary analysis endpoints were 
(1) to compare demographic, socioeconomic and clinical characteristics between the mental health clusters of mountain accident victims and 
(2) to identify patients at risk of post-traumatic mental disorders based on explanatory variables available during early medical treatment of the mountain sport accident.

@@

The study collective consisted of victims of mountain sport accidents which happened in Tyrol, Austria, a renowned and popular alpine touristic region. 
In- and outpatient adult survivors of mountain sport accidents and treated at the Department of Orthopedics and Traumatology at Medical University of Innsbruck (Tyrol, Austria) between 1^st^ January 2018 and 31^st^ December 2020 were invited to participate (n = `r nrow(ptsd$cleared)`). 
Among the invited patients, 
`r length(ptsd$interview_ID)` completed the study survey and 
`r nrow(ptsd$dataset)` individuals with the complete psychometric data set were analyzed (overall analysis inclusion rate: 
`r ((nrow(ptsd$dataset)/nrow(ptsd$cleared)) * 100) %>% signif(2)`%, __Figure \@ref(fig:fig-consort)__, __Supplementary Tables S1__ - __S3__). 
The median time between the trauma center admission and survey completion was 
`r median(ptsd$dataset$obs_time)` days (interquartile range: `r quantile(ptsd$dataset$obs_time, c(0.25, 0.75)) %>% paste(collapse = ' - ')` days). 
As compared with the analyzed participants, patients who did not respond to the study invitation were characterized by a higher frequency of biking accidents, lower rate of alpine ski/snowboard accidents, less severe injury, lower hospitalization and surgery rates, and less surgery diagnoses (__Supplementary Table S4__). 
The survey participants excluded from the analysis due to missingness of psychometric data had significantly lower income, less severe injuries, more frequent injuries of the upper limbs, less frequently required hospitalization or surgery, and suffered less frequently from persistent physical health consequences of the accident than the analyzed participants (__Supplementary Table S5__). 
Effect sizes of the differences between the included and excluded participants were weak (V $\leq$ 0.22 or r $\leq$ 0.15).

The analyzed participants were predominantly middle-aged 
(median: `r get_num_stats(ptsd$dataset, 'age', collapse = ', interquartile range: ')` years) and 
`r get_percent(ptsd$dataset, 'sex')$percent['female']`% of them were females.
The vast majority had secondary or tertiary education grade 
(`r sum(get_percent(ptsd$dataset, 'education')$percent[c('secondary', 'tertiary')])`%) and were professionally active 
(`r get_percent(ptsd$dataset, 'employment_status')$percent['employed']`%). 
Less than 8% of participants worked in a trauma-exposed or mountain sport profession. 
High annual household incomes of $\geq$ 45000 Euro was reported by 
`r get_percent(ptsd$dataset, 'high_income')$percent['yes']`% of participants. 
Less than 10% of participants were smokers or at risk of problematic alcohol use. 
Pre-existing physical disorders were reported by 
`r get_percent(ptsd$dataset, 'somatic_comorbidity')$percent['yes']`% of participants with cardiovascular, neurological and metabolic illness being the most frequent. 
Four of ten participants had experiences by or witnessed a traumatic event prior to the accident, 
`r sum(get_percent(ptsd$dataset, 'traumatic_number')$percent[c('2', '3+')])`% participants experienced two or more traumatic events. 
Mental disorders prior the accident affected 
`r get_percent(ptsd$dataset, 'psych_comorbidity')$percent['yes']`% of the cohort with affective 
(`r get_percent(mental$analysis_tbl, 'affective')$percent['yes']`%) and somatoform disorders 
(`r get_percent(mental$analysis_tbl, 'somatoform')$percent['yes']`%) as the leading pre-existing mental conditions (__Table \@ref(tab:tab-cohort-demo)__).

Mountain sport accidents in the past were reported by `r get_percent(ptsd$dataset, 'prior_accident')$percent['yes']`% of participants. 
Most of the investigated accidents occurred during alpine skiing, snowboarding or cross-country skiing 
(`r get_percent(ptsd$dataset, 'sport_type')$percent['ski/snowboard/cross-country']`%) followed by biking 
(`r get_percent(ptsd$dataset, 'sport_type')$percent['biking']`%) and classical mountain sports such as climbing, hiking, mountaineering or back-country skiing 
(`r get_percent(ptsd$dataset, 'climbing/hiking/mountaineering/skitour')$percent['climbing/hiking/mountaineering/skitour']`%). 
One-third of participants were alone during the accident 
(`r get_percent(ptsd$dataset, 'accident_alone')$percent['yes']`%) and, in most cases, were the sole culprit 
(`r get_percent(ptsd$dataset, 'accident_culprit')$percent['self']`). 
Professional rescue service was involved in 
`r get_percent(ptsd$dataset, 'accident_rescue')$percent['rescue team']`% of the accidents. 
In `r get_percent(ptsd$dataset, 'injury_sev_strata')$percent['2']`% of participants the injury was moderate (AIS 2) and in 
`r get_percent(ptsd$dataset, 'injury_sev_strata')$percent['3+']`% severe-to-critical (AIS $\geq$ 3). 
Limb injuries were the most common followed by head and face (__Supplementary Figure S1A__). 
Hospitalization and surgery rates following the accident were 
`r get_percent(ptsd$dataset, 'hospitalization')$percent['yes']`% and 
`r get_percent(ptsd$dataset, 'surgery_done')$percent['yes']`%, respectively. 
Psychological or psychiatric support after the accident was provided to 
`r get_percent(ptsd$dataset, 'psych_support_post_accident')$percent['yes']`% individuals. 
A subset of participants (`get_percent(ptsd$dataset, 'psych_support_need')$percent['yes']`%), who had not received it, declared need for psychological support following the accident. 
Despite persisting physical consequences of the accident, which affected 
`r get_percent(ptsd$dataset, 'accident_aftermath')$percent['yes']`% of participants and flashbacks during mountain sport at the time of the questionnaire completion in 
`r get_percent(ptsd$dataset, 'unwilling_flashback')$percent['yes']`% of the cohort, 
most participants returned to the same mountain sport following the accident 
(`r get_percent(ptsd$dataset, 'same_sport_type_post_accident')$percent['yes']`%). 
Yet, `r get_percent(ptsd$dataset, 'caution_post_accident')$percent['more cautious']`% described their behavior during mountain sport as more cautious (__Table \@ref(tab:tab-cohort-accident)__).

>>>>

author: 
  - Hanna Salvotti, MD:
      institute: [psych, regensburg]
  - Piotr Tymoszuk, PhD:
      institute: [daas]
  - Mathias Ströhle, MD:
      institute: [kufstein, himed]
  - Peter Paal, MD: 
      institute: [himed, salzburg, kurasi, icar]
  - Hermann Brugger, MD: 
      institute: [icar. eurac, ismm]
  - Martin Faulhaber, PhD: 
      institute: [himed, sport]
  - Nicola Kruijer, MD: 
      institute: [trauma]
  - Barbara Sperner-Unterweger, MD: 
      institute: [psych]
  - Katharina Hüfner, MD:
      institute: [psych, himed]
      email: katharina.huefner@tirol-kliniken.at
      correspondence: true

institute:
  - psych: Department of Psychiatry, Psychotherapy, Psychosomatics and Medical Psychology, University Hospital for Psychiatry II, Medical University of Innsbruck, Innsbruck, Austria.
  - regensburg: Department of Neurosurgery, University Hospital of Regensburg, Regensburg, Germany
  - daas: Data Analysis as a Service Tirol, Wörgl, Austria
  - kufstein: Department of Anesthesiology and Critical Care Medicine, Bezirkskrankenhaus Kufstein, Kufstein, Austria.
  - himed: Austrian Society of Mountain and High-Altitude Medicine, Mieming, Austria. 
  - salzburg: Department of Anesthesiology and Critial Care Medicine, Paracelsus Medical University, Salzburg, Austria.
  - kurasi:  Austrian Board of Mountain Safety (Österreichisches Kuratorium fur Alpine Sicherheit), Innsbruck, Austria.
  - icar: International Commission for Mountain Emergency Medicine (ICAR), Kloten, Switzerland.
  - eurac: EURAC Research, Institute for Mountain Emergency Medicine, Bolzano/Bozen, Italy.
  - ismm: International Society of Mountain Medicine (ISMM).
  - sport: Department of Sport Science, University of Innsbruck, Innsbruck, Austria.
  - trauma: Department of Orthopedics and Traumatology, Medical University of Innsbruck, Innsbruck, Austria.

@@

__Background:__ The restorative effect of physical activity in natural alpine environments on mental and physical health is increasingly recognized. 
However, mountain sports are also linked to a risk of accidents. 
We aimed to characterize mental health in individuals following mountain sport accidents requiring professional medical management.

__Methods:__ Adult victims of mountain sport accidents treated at the hospital of the Medical University of Innsbruck (Austria) between 2018 and 2020 completed a cross-sectional survey at a median `r round(median(ptsd$dataset$obs_time)/30.436)` months following the admission (n = `r nrow(ptsd$dataset)`). 
Symptoms of post-traumatic stress disorder (PTSD, PCL-5), anxiety, depression, and somatization (PHQ), resilience (RS-13), sense of coherence (SOC-9L), post-traumatic growth (PTGI) and quality of life (EUROHIS-QOL), as well as sociodemographic and clinical information were obtained from an online survey and extracted from electronic health records. 
Mental health outcome patterns were investigated by semi-supervised medoid clustering and modeled by machine learning.

__Results:__ Symptoms of PTSD were observed in `r signif(table(ptsd$dataset$dsm5_cluster_class)[2]/nrow(ptsd$dataset), 2) * 100`% of participants. 
Three comparably sized subsets of participants were identified: a (1) neutral, (2) post-traumatic growth and (3) post-traumatic stress cluster. 
The post-traumatic stress cluster was characterized by high prevalence of symptoms of mental disorder, low resilience, low sense of coherence and low quality of life as well as by younger age, the highest frequency of pre-existing mental disorders and persisting physical health consequences of the accident. Individuals in this cluster self-reported a  need for psychological support following the accident and more cautious behavior during mountain sports since the accident. 
Reliability of machine learning-based prediction of the cluster assignment based on `r length(class_globals$variables$early)` variables available during acute medical treatment of accident victims was limited.

__Conclusion:__ A subset of individuals develop symptoms of mental health disorders including symptoms of PTSD when assessed at least 6 months after the mountain sport accident. 
Since early identification of these vulnerable patients remains challenging, psychoeducational measures for all patients and low-threshold access to psychological support is key for a successful interdisciplinary management of victims of mountain sport accidents.

@@

mountain sport accident, machine learning, mental health, post-traumatic growth, post-traumatic stress disorder

@@

Physical activity in mountains, such as hiking, climbing, alpine skiing or ski touring, are gaining increasing popularity, a trend that has been accelerated by the COVID-19 pandemic [@Hedenborg2022]. 
Outdoor physical activity in general and mountain sports in particular exert beneficial effects on physical and mental health [@Ower2019; @Niedermeier2017a; @Niedermeier2017]. 
However, mountain sports are also associated with the risk of accidents and injury [@Faulhaber2017; @Faulhaber2020], which could consequently lead to development of post-traumatic stress disorder (PTSD) or other stress-related mental disorders.

Diagnostic symptoms of PTSD  include re-experiencing (criterion B), avoidance (criterion C), negative alterations in cognition and mood (criterion D) and hyper-arousal (criterion E) [@AmericanPsychiatricAssociation2013], which can occur following self-experiencing or witnessing of a traumatic event such as actual or threatened death, injury, physical or sexual violence. 
PTSD may co-occur with symptoms of depression and anxiety and substance abuse disorders [@Gagnon-Sanschagrin2022; @Brady2000]. 
Although most individuals have experienced at least one traumatic event during their lifetime [@Benjet2016], the prevalence of PTSD in the general European population has been estimated between 0.38 to 6.67% [@Burri2014]. 
Yet, up to 26% of acute physical trauma patients have been reported to develop PTSD [@Joseph2020].

So far it has proven difficult to reliably identify early predictors of PTSD development in accident victims. 
While some recent studies on specific populations have identified risk factors, such as pre-existing psychological distress [@Ziobrowski2021] or symptoms of mental disorders [@Papini2018], others have focused on the arousal and stress reaction during acute trauma and peri-traumatic period as candidate predictors [@Schultebraucks2020]. 
These results require validation and further generalization before they can be implemented into routine care [@Wafa2019; @Shalev2019].

Resilience, i.e. the ability to recover from adversities can protect an individual from developing PTSD and can help to explain why most people do not develop PTSD despite the high prevalence of traumatic events in the general population [@Ahmed2007; @Bensimon2012]. 
Generally individuals who perform mountain sports display high resilience values [@Ower2019]. 
Post-traumatic growth can occur in individuals struggling to overcome the consequences of a traumatic event and describes a psychological adjustment reaction with strengthened positive perceptions of self, others, one's life, and meaning of the event [@Tedeschi1996]. 
Although both resilience and post-traumatic growth are salutogenetic concepts their exact link is still under investigation [@Levine2009].

With the increasing popularity of mountain sports, it is essential to characterize the mental health consequences of accidents during mountain sports. 
Prevention of PTSD or other mental health problems following an accident is essential to preserve the positive mental health effects of physical activity in an mountain environment despite the accident risk. 
Available data on the consequences of accidents during mountain sports on mental health are scarce. 
Prior studies were performed on mountain guides, rescuers or avalanche accident survivors and mostly lack clinical data [@Brugger2002; @Mikutta2022; @Traber2020; @Leonard2021; @Thordardottir2015; @Sommer2004]. 
We analyzed mental health readouts such as symptoms of PTSD, anxiety, depression, and somatization, resilience, sense of coherence, post-traumatic growth and quality of life in `r nrow(ptsd$dataset)` individuals treated at the Hospital of the Medical University of Innsbruck (Austria) after a mountain sport accident by means of semi-supervised clustering. 
Additionally, we searched for early sociodemographic and clinical predictors of mental health impairment following a mountain sport accident with a machine learning approach.

@@

Individuals treated following a mountain sport accident at the Department for Orthopedics and Traumatology at the University Clinic of Innsbruck between 1^st^ January 2018 and 31^st^ December 2020 were screened for participation. 
Individuals fulfilling the inclusion criteria: 
hospital admission at least 6 months prior to the start of the study, 
age $\geq$ 18 years, residence in a German-speaking country, and 
proficiency in German (n = `r nrow(ptsd$cleared)`) were invited to participate in the online study survey via conventional mail. 
Out of the invited subjects, `r length(ptsd$interview_ID)` completed the survey. 
Surveys of `r nrow(ptsd$dataset)` individuals with the complete psychometry data were analyzed (overall response rate: `r signif(nrow(ptsd$dataset)/nrow(ptsd$cleared), 2) * 100`%, __Figure \@ref(fig:fig-consort)__, __Supplementary Tables S1__ - __S3__).

@@

Sociodemographic (e.g. age, sex, education, profession, income, prior mountain sport accidents and prior traumatic events), accident- and recovery-related (e.g. mode of rescue, psychological support, persistent physical health consequences) and psychometric variables were recorded with a cross-sectional online survey.

The psychometric battery consisted of German versions of assessment tools for symptoms of 
PTSD (PCL-5: PTSD checklist for DSM-5) [@Bovin2016], 
anxiety (GAD-7: 7-item general anxiety disorder scale) [@Spitzer2006], 
depression (PHQ-9: 9-item patient health questionnaire for depressive symptoms) [@Grafe2004], 
panic (PHQ-panic module) [@Grafe2004], 
somatization (PHQ-15: 15-item patient health questionnaire of common somatic symptoms as a substitute for somatization) [@Kroenke2002], 
resilience (RS13: 13-item resilience scale) [@Leppert2008], 
sense of coherence (SOC-9L: Leipzig 9-item sense of coherence questionnaire) [@Schumacher2000], 
quality of life (EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale) [@Schmidt2006], 
and post-traumatic growth (PTGI: post-traumatic growth inventory) [@Tedeschi1996]. 
The tools displayed good-to-excellent consistency [@McDonald1999] (McDonald's $\omega$ > 0.8, __Supplementary Table S2__ and __S3__).

Clinically significant symptoms of anxiety (GAD-7 $\geq$ 10), 
depression (PHQ-9 $\geq$ 11) [@Manea2012], 
somatization (PHQ-15 $\geq$ 11) [@Kroenke2002] and 
resilience classes (low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73) [@Leppert2008] 
were defined with published cutoffs.

Separate scores were calculated for domains B, C, D and E of the PCL-5 tool along with the total score being the sum of all items. 
Each PCL-5 item was scored as 0: not at all, 1: a little bit, 2: moderate, 3: quite a bit and 4: extremely. 
Participants positive for the B domain or C domain PTSD symptoms were identified by at least one item per domain scored with 'moderate' or higher. 
Participants positive for the D or E domain PTSD symptoms were identified by at least two items per domain scored with 'moderate' or higher. 
Significant PTSD symptoms were assumed in participants screened positive for at least one of the B, C, D or E PCL-5 domains [@Bovin2016].

Traumatic events prior to the accident were assessed with the DIA-X tool [@Maercker2005]. 
Flashback frequency was surveyed as none, > 1/year and > 1/month. 
Smoking was surveyed as a single yes/no question. 
Problematic alcohol use was investigated with the CAGE tool [@OBrien2008]. 
Data on the type of the mountain sport activity during the accident, injury diagnosis, severity (AIS: abbreviated injury scale) [@Gennarelli2006] and body location, hospitalization, surgery and number of surgical ICD-10 diagnoses were extracted from electronic patient records.

@@

The primary analysis endpoint was assessment of mental health following a mountain sport accident by clustering using standardized psychometric tools for symptoms of PTSD, anxiety, depression, panic, somatization, resilience, sense of coherence, post-traumatic growth, and quality of life. 
The secondary analysis endpoints were 
(1) to compare demographic, socioeconomic and clinical characteristics between the different mental health clusters of mountain accident victims and 
(2) to identify patients at risk of mental disorders based on a set of explanatory variables available during acute medical treatment after mountain sport accidents.

@@

Surveys with complete psychometric data from `r nrow(ptsd$dataset)` individuals were analyzed (__Figure \@ref(fig:fig-consort)__). 
The median time between the trauma center admission after the accident and the survey completion was 
`r median(ptsd$dataset$obs_time)` days (interquartile range: `r quantile(ptsd$dataset$obs_time, c(0.25, 0.75)) %>% paste(collapse = ' - ')` days). 
As compared with the analyzed participants, patients who did not respond to the study invitation were characterized by a higher frequency of biking accidents, lower rate of alpine ski/snowboard accidents, less severe injury, lower hospitalization and surgery rates, and fewer surgery diagnoses (__Supplementary Table S4__). 
The survey participants excluded from the analysis due to missingness of psychometric data had significantly lower income, less severe injuries, more frequent injuries of the upper limbs, less frequently required hospitalization or surgery, and suffered less frequently from persistent physical health consequences of the accident than the analyzed participants (__Supplementary Table S5__). 
Effect sizes of the differences between the included and excluded participants were weak (V $\leq$ 0.22 or r $\leq$ 0.15).

The analyzed participants were predominantly middle-aged 
(median: `r get_num_stats(ptsd$dataset, 'age', collapse = ', interquartile range: ')` years) and 
`r get_percent(ptsd$dataset, 'sex')$percent['female']`% of them were females.
The vast majority had secondary or tertiary education grade 
(`r sum(get_percent(ptsd$dataset, 'education')$percent[c('secondary', 'tertiary')])`%) and were professionally active 
(`r get_percent(ptsd$dataset, 'employment_status')$percent['employed']`%). 
Less than 8% of participants worked in a search and rescue or mountain sport profession. 
Annual household incomes of $\geq$ 45000 Euro were reported by 
`r get_percent(ptsd$dataset, 'high_income')$percent['yes']`% of participants. 
Less than 10% of participants were smokers or at risk of problematic alcohol use defined as CAGE $\geq$ 2. 
Pre-existing physical disorders were reported by 
`r get_percent(ptsd$dataset, 'somatic_comorbidity')$percent['yes']`% of participants with cardiovascular, neurological and metabolic illness being the most frequent. 
Mental disorders diagnosed by a physician before the accident affected 
`r get_percent(ptsd$dataset, 'psych_comorbidity')$percent['yes']`% of the cohort with affective 
(`r get_percent(mental$analysis_tbl, 'affective')$percent['yes']`%) and somatoform disorders 
(`r get_percent(mental$analysis_tbl, 'somatoform')$percent['yes']`%) as the leading pre-existing mental conditions. 
Four of ten participants had experienced by or witnessed a traumatic event prior to the accident, 
`r sum(get_percent(ptsd$dataset, 'traumatic_number')$percent[c('2', '3+')])`% participants had been exposed to two or more traumatic events (__Table \@ref(tab:tab-cohort-demo)__).

Mountain sport accidents in the past were reported by `r get_percent(ptsd$dataset, 'prior_accident')$percent['yes']`% of participants. 
Most of the investigated accidents occurred during alpine skiing, snowboarding or cross-country skiing 
(`r get_percent(ptsd$dataset, 'sport_type')$percent['ski/snowboard/cross-country']`%) followed by biking 
(`r get_percent(ptsd$dataset, 'sport_type')$percent['biking']`%) and classical mountain sports such as climbing, hiking, mountaineering or back-country skiing 
(`r get_percent(ptsd$dataset, 'sport_type')$percent['climbing/hiking/mountaineering/skitour']`%). 
One-third of participants were alone during the accident 
(`r get_percent(ptsd$dataset, 'accident_alone')$percent['yes']`%) and, in most cases, were the only person responsible to the accident 
(`r get_percent(ptsd$dataset, 'accident_culprit')$percent['self']`). 
Professional rescue service was involved in 
`r get_percent(ptsd$dataset, 'accident_rescue')$percent['rescue team']`% of the accidents. 
In `r get_percent(ptsd$dataset, 'injury_sev_strata')$percent['2']`% of participants the injury severity was moderate (AIS 2) and in 
`r get_percent(ptsd$dataset, 'injury_sev_strata')$percent['3+']`% severe-to-critical (AIS $\geq$ 3). 
Limb injuries were the most common followed by injuries of the head and face (__Supplementary Figure S1A__). 
Hospitalization and surgery rates were 
`r get_percent(ptsd$dataset, 'hospitalization')$percent['yes']`% and 
`r get_percent(ptsd$dataset, 'surgery_done')$percent['yes']`%, respectively. 
Psychological or psychiatric support after the accident was provided to 
`r get_percent(ptsd$dataset, 'psych_support_post_accident')$percent['yes']`% individuals. 
A subset of participants (`r get_percent(ptsd$dataset, 'psych_support_need')$percent['yes']`%), who had not received psychological or psychiatric support, declared a need for psychological or psychiatric support following the accident. 
Persisting physical consequences of the accident were reported by 
`r get_percent(ptsd$dataset, 'accident_aftermath')$percent['yes']`% of participants and flashbacks during mountain sport at the time of the questionnaire completion were observed in 
`r get_percent(ptsd$dataset, 'unwilling_flashback')$percent['yes']`% of the cohort. 
Although most individuals returned to the same mountain sport following the accident 
(`r get_percent(ptsd$dataset, 'same_sport_type_post_accident')$percent['yes']`%), 
`r get_percent(ptsd$dataset, 'caution_post_accident')$percent['more cautious']`% of all participants described their behavior during mountain sport as more cautious (__Table \@ref(tab:tab-cohort-accident)__).

At least one diagnostic criterion of PTSD defined as positive scoring of the B, C, D or E domains of the PCL-5 tool was met by 
`r get_percent(ptsd$dataset, 'dsm5_cluster_class')$percent['positive']`% of participants. 
The domain B PTSD symptoms were the most common (`r get_percent(ptsd$dataset, 'dsm5_B_class')$percent['positive']`%) and the domain D symptoms were the least frequent (`r get_percent(ptsd$dataset, 'dsm5_D_class')$percent['positive']`%). 
Solely `r digit2string(sum(reduce(cohort$ptsd_overlap$data, function(x, y) x * y) > 0))` patients 
(`r signif(sum(reduce(cohort$ptsd_overlap$data, function(x, y) x * y) > 0)/nrow(cohort$ptsd_overlap$data), 2) * 100`%) were screened positive for all four diagnostic criteria of PTSD (__Table \@ref(tab:tab-cohort-mental)__, __Supplementary Figure S1BC__). 
Clinically relevant symptoms of 
anxiety (`r get_percent(ptsd$dataset, 'gad7_total_class')$percent['positive']`%), 
depression (`r get_percent(ptsd$dataset, 'phq9_total_class')$percent['positive']`%) 
and somatization (`r get_percent(ptsd$dataset, 'phq_events_total_class')$percent['positive']`%) were rare in this cohort consisting of `r get_percent(ptsd$dataset, 'rs13_total_class')$percent['high']`% highly resilient individuals (__Table \@ref(tab:tab-cohort-mental)__). 
We could not observe any significant association of values of psychometric scores or frequency of mental health problems with the time between trauma center admission and survey completion (not shown).
